Episodic memory: mental time travel or a quantum `memory wave' function?

### 2020-09-21 paper/boneyard.tex
working on qmwf vs mtt section
finished draft of section on word embedding models

>>>>

A nuance of the quantum memory wave function framework is that,

However, because the thoughts we have at a given moment may themselves draw on other moments of our past,

The central limitation of the mental time travel metaphor that the quantum memory wave function addresses is that

  .  The transformed outputs distort the original inputs in a way that often enables subseqeuent components of a deep text embedding model to compute more stable and meaningful embeddings.  \textit{Deep averaging networks} pool information from different observations in the original data in order to magnify small but meaningful differences.  \textit{Recurrent neural networks} (and variants) are often used to model temporal relations in sequence data~\citep[the deep averaging network and recurrent neural network diagrams are loosely inspired by figures from][]{IyyeEtal15}.  \textit{Deep autoencoders} attempt to reconstruct inputted patterns after passing them through a series of increasingly smaller layers in an encoding network.  The resulting activation patterns in the discriminator layer, which contains the fewest number of neurons, may be used as a low-dimensional embedding of the input.  The discriminator's activations are then passed through a decoder network which contains successively larger layers.  Finally, a reconstructed output is produced.  After fitting this model, outputs may be estimated from arbitrary activations of the discriminator layer~\citep[diagram loosly inspired by][]{YousHame17}.

### 2020-09-21 paper/main.tex
finished draft of section on word embedding models

\subsection*{Defining the geometries of ongoing thoughts and memories}
Defining geometric spaces that can appropriately capture the conceptual contents of our thoughts and experiences has become a major focus of computational science.  An class of models that have shown particular promise over the past several decades is broadly referred to as \textit{word embedding models} (Fig.~\ref{fig:embedding-models}).  The fundamental goal of these models is to take as input a sequence of words (text) and produce as output one or more feature vectors whose coordinates reflect their conceptual (semantic) content.  For example, when two similar documents are passed through these models, the goal is to produce two feature vectors that are closer (e.g., in Euclidean distance) than the feature vectors for two dissimilar documents.

Early word embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{LandDuma97} attempted to compute these embeddings using word co-occurance statistics.  The logic driving these approaches posited that documents that contained similar mixtures of words would be likely to be conceptually similar, and words that appeared in similar documents would also likely be conceptually similar.  To quantify these intuitions, \cite{LandDuma97} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of time a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents by number-of-words word count matrix into the product of a number-of-documents $\times K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K \times $ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflected the similarities between documents, and the columns of the word embeddings matrix reflected similarities between words.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.7\textwidth]{figs/word_embedding_models}
\caption{\textbf{Three approaches to text embedding.}  \textbf{A. Matrix factorization models.} Per-document (or per-participant) word counts and/or frequencies are collected into a word count matrix (left).  To compute $K$-dimensional embeddings of the documents and words, the word count matrix is factorized (typically by computing its Singular Value Decomposition) into the product of a number-of-documents $\times K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K \times $ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflect the similarities between documents, and the columns of the word embeddings matrix reflect similarities between words.  \textbf{B. Topic models.}  Fitting a topic model to a word count matrix (e.g., Panel A) reveals a set of latent \textit{topics}, or themes that pervade each document.  The topics are specified in a $K \times$ number-of-words \textit{topics matrix}, and the per-document themes are specified in the number-of-documents $\times K$ \textit{topic proportions matrix}. A given document's topic proportions reflect the specific mix of themes that are reflected in that document (bar graph on the left).  These proportions are discovered automatically by estimating the topic from which each word in the document is most probably drawn (highlighted words in \textit{Document text}).  \textbf{C. Deep learning models.}  A wide variety of deep learning approaches to text embedding have been proposed.  Broadly, these approaches often combine a common set of elements.  \textit{Transformer} models take a sequence as input and output a new modified sequence~\citep[diagram adapted from][]{ViswEtal17}.  The transformed outputs distort the original inputs in a way that often enables subseqeuent components of a deep text embedding model to compute more stable and meaningful embeddings.  \textit{Deep averaging networks} pool information from different observations in the original data in order to magnify small but meaningful differences.  \textit{Recurrent neural networks} (and variants) are often used to model temporal relations in sequence data~\citep[the deep averaging network and recurrent neural network diagrams are loosely inspired by figures from][]{IyyeEtal15}.  \textit{Deep autoencoders} attempt to reconstruct inputted patterns after passing them through a series of increasingly smaller layers in an encoding network.  The resulting activation patterns in the discriminator layer, which contains the fewest number of neurons, may be used as a low-dimensional embedding of the input.  The discriminator's activations are then passed through a decoder network which contains successively larger layers.  Finally, a reconstructed output is produced.  After fitting this model, outputs may be estimated from arbitrary activations of the discriminator layer~\citep[diagram loosly inspired by][]{YousHame17}.}
\label{fig:embedding-models}
\end{figure}

@@

% deep learning approaches
% bag of words vs. sequence models, context and grammar to convey meaning, etc.

>>>>

\subsection*{Defining the geometries of ongoing thoughts and memories}
Defining geometric spaces that can appropriately capture the conceptual contents of our thoughts and experiences has become a major focus of computational science.  One family of models that has shown particular promise over the past several decades is broadly referred to as \textit{word embedding models} (Fig.~\ref{fig:embedding-models}).  The fundamental goal of these models is to take as input a sequence of words (text) and produce as output one or more feature vectors whose coordinates reflect their conceptual (semantic) content.  For example, when two similar documents are passed through these models, the goal is to produce two feature vectors that are closer (e.g., in Euclidean distance) than the feature vectors for two dissimilar documents.

Early word embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{DeerEtal90, LandDuma97} attempted to compute these embeddings using word co-occurance statistics.  The logic driving these approaches posited that documents that contained similar mixtures of words would be likely to be conceptually similar, and words that appeared in similar documents would also likely be conceptually similar.  To quantify these intuitions, \cite{LandDuma97} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of time a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents by number-of-words word count matrix into the product of a number-of-documents $\times K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K \times $ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflected the similarities between documents, and the columns of the word embeddings matrix reflected similarities between words.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.75\textwidth]{figs/word_embedding_models}
\caption{\textbf{Three approaches to text embedding.}  \textbf{A.~Matrix factorization models.} Per-document (or per-participant) word counts and/or frequencies are collected into a word count matrix (left).  To compute $K$-dimensional embeddings of the documents and words, the word count matrix is factorized (typically by computing its Singular Value Decomposition) into the product of a number-of-documents $\times K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K \times $ number-of-unique-words \textit{word embeddings matrix}.  \textbf{B.~Topic models.}  Fitting a topic model to a word count matrix (e.g., Panel A) reveals a set of latent \textit{topics}, or themes that pervade each document.  The topics are specified in a $K \times$ number-of-words \textit{topics matrix}, and the per-document themes are specified in the number-of-documents $\times K$ \textit{topic proportions matrix}. A given document's topic proportions reflect the specific mix of themes that are reflected in that document (bar graph on the left).  These proportions are discovered automatically by estimating the topic from which each word in the document is most probably drawn (highlighted words in \textit{Document text}).  \textbf{C.~Deep learning models.}  A wide variety of deep learning approaches to text embedding have been proposed.  Broadly, these approaches often combine a common set of elements, including \textit{transformers}~\citep[diagram adapted from][]{ViswEtal17}, \textit{deep averaging networks}~\citep[diagram adapted from][]{IyyeEtal15}, \textit{recurrent neural networks}~\citep[diagram adapted from][]{IyyeEtal15}, and \textit{deep autoencoders}~\citep[diagram adapted from][]{YousHame17}.}
\label{fig:embedding-models}
\end{figure}

@@

Both the matrix factorization models and topic model described above are each examples of \textit{bag of words} models.  Given a document, these models will generate the same embeddings for any random re-ordering of the same words.  This over-simplifies the intended conceptual content of documents, where word order often conveys important meaning.  An alternative framing of the text embedding problem is to use \textit{sequence} models that explicitly account for word order effects.  Most modern sequence-based models are implemented using deep neural networks (Fig.~\ref{fig:embedding-models}C).

The Neural Probabilistic Language Model~\citep[NPLM;][]{BengEtal03} was one of the earliest deep learning-based sequence models.  NLPM derives word embeddings for each word in the vocabulary (such that nearby embeddings reflect semantically related words), and then models word sequences according to their meanings.  This model was inspired in part by $n$-gram models~\citep{BrowEtal92}, which estimate the probability of the next word in a sequence given the $n$ words that came before it. Effectively, NLPM attempts to learn arbitrarily long sequences.  In a related approach, \cite{MnihHint09} proposed the Scalable Hierarchical Distributed Language Model, which learns sequences of $n$-grams to achieve greater explanatory power.

The next major advance in word embeddings came from attempts to more precisely capture the geometric relations between words.  For example, if each word's semantic meaning is captured by a feature vector, then in principle one could perform vector algebra on different words' embeddings to produce new concepts that combine their meanings.  For example, subtracting the vector for ``man'' from that of ``king'' and then adding the vector for ``woman'' would result in a new feature vector similar to that of ``queen''~\citep{MikoEtal13b}.  \cite{MikoEtal13a} developed the deep learning model \textit{word2vec} to explicitly optimize the accuracy of these vector operations.  Whereas word2vec is a bag of words model, Global Vectors for Word Representation~\citep[GloVe;][]{PennEtal14} extends this approach to model each word's local (temporal) context~\citep[also see][]{PeteEtal18}.

Several recent models have begun to move beyond single word-level embeddings.  For example, SkipThought~\citep{KiroEtal15}, Universal Sentence Encoder~\citep{CerEtal18}, InferSent~\citep{ConnEtal18}, and RandSent~\citep{WietKiel19} each compute sentence-level embeddings.  The central insight of these approaches is that capturing the nuanced meaning of a particular concept often requires \textit{several} words that have been organized into a complete sentence or sentence-like grammatical construct.

Cutting-edge approaches to word embeddings that have achieved the best-to-date performance on modern natural language processing benchmarks are built using enormous deep neural networks often with many billions of parameters.  Because these models are very computationally expensive to fit, an important discovery has been that many layers in these models may be pre-trained without substantially affecting performance, even on new tasks.  One of the first of this new class of model, Bidirectional Encoder Representations from Transformers~\citep[BERT;][]{DevlEtal18}, uses a pre-trained deep neural network, whose weights in a single output layer may be fine-tuned to suit a wide range of new tasks.  The latest additions to this class of models are called Generative Pre-Trained (GPT) transformers~\citep{RadfEtal18, RadfEtal19, BrowEtal20}.

% deep learning approaches:
% - bag of words vs. sequence models vs. sentence-level embeddings
% - accounting for use of context and grammar to convey meaning
% - pre-training vs. customized training vs. training-free

% Neural Probabilistilc Language Model (NPLM): BengEtal03 --> explicitly model word sequences
% Scalable Hierarchical Distributed Language Model (SHDLM): MnihHint09 --> sequences of word trees

%word2vec: MikoEtal13 --> semantic relations reflected in geometric operations
%GloVe: PennEtal14 --> combines matrix factorization and local context window methods
%SkipThought: KiroEtal15 --> reconstructs surrounding sentences of an encoded passage
%BERT: DevlEtal18 --> pretraining language representations, context dependence of word embeddings
%ELMO: PeteEtal18 --> how is it different from BERT/GPT?
%InferSent: ConnEtal18 --> embeddings of sentences
%Universal Sentence Encoder: CerEtal18 --> embeddings of sentences
%Open-GPT: RadfEtal18 --> pretrained transformer (how is it different from BERT?)
%RandSent: WietKiel19 --> training-free sentence-level embeddings
%GPT-2: RadfEtal19
%GPT-3: BrowEtal20
